Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers

### 2019-09-12 Manuscript/manuscriptPS.tex

lipid/counter-ion & force field & \footnote{Number of lipid molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Number of reference for simulation files}files & \tabularnewline

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with membrane in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were calculated by first averaging over time separately
for each lipid in the system, and then calculating the average and
the standard error of the mean over the different lipids. The analysis can be conducted using the
Python program {\tt calcOrderParameters.py}, available in Ref.~\citenum{MATCHgit}, that uses the
MDAnalysis library \cite{agrawal11,gowers16}.
For the united atom force fields, the positions of hydrogens were generated before the order parameter calculation using the {\tt protonate} tool
of the Gromacs 3 sofware package \cite{gromacsMANUAL}.
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

\subsection{Using the molecular electrometer to compare ion binding to negatively charged lipid bilayers 
in simulations and in experiments}

The $S_{\rm CH}$ of the $\alpha$ and $\beta$ carbons in the PC headgroup
decrease proportionally to the amount of positive
charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
and can therefore be used to measure the ion binding affinity.
In addition to ions, the correlation between bound charge and headgroup order parameter change
is empirically observed also for peptides, charged amphiphiles, local anesthetics and charged lipids~\cite{scherer87,beschiaschvili91}.
This concept, known as the molecular electrometer, is especially useful for 
comparison between simulations and experiments, as
the headgroup $S_{\rm CH}$ at varying cation
concentrations can be easily calculated from
simulations~\cite{catte16}. The headgroup $S_{\rm CH}$
of negatively charged PS and PG lipids also exhibit systematic dependencies on the bound charge, but these are less well understood than for PC~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, measuring the PC headgroup $S_{\rm CH}$ in 
mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also Supporting Information section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing ion binding to such negatively charged membranes.

Calibrating the PC headgroup $S_{\rm CH}$ response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer.
This can be done using experimental data from mixtures of PC and
monovalent cationic surfactants (such as POPC and dihexadecyldimethylammonium,
see SI section \ref{electrometerCALIBRATION})~\cite{scherer89,melcr18}.
Additionally, the response of PC headgroup $S_{\rm CH}$ to the negatively
charged PS follows the molecular electrometer in experiments~\cite{scherer87},
which we also quantify here (see SI section \ref{electrometerFORmixtures}).

Studies applying the molecular electrometer have used two different definitions for salt concentration:
The concentrations reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids.
In the former case, binding of ions to the lipids leads to a lower bulk concentration than what was present in the original solvent.
However, the choice of definition has only a marginal effect
to the results in simulations with realistic ion binding affinity
(see SI section \ref{concentrationDEFsection}).
In this work, we use the concentration before solvating the lipids to be consistent with the reference
experimental data \cite{roux90}.

\section{Results and Discussion}

\subsection{Headgroup and glycerol backbone C--H bond order parameters of POPS from $^{13}$C NMR}
The INEPT and 2D R-PDLF experiments from POPS samples gave well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}, whereas
the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
known C--H bond order parameters from the $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
are shown in the Supporting Information (Fig. \ref{DPslices}).

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) Refocused-INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental S-DROSS data (points), and SIMPSON simulations (blue lines) with
    the C--H bond order parameter values of $-0.12$ for the $\beta$-carbon, and $+0.09$ and $-0.02$
    for the $\alpha$-carbon.
    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value ($+0.02$) 
    for the smaller $\alpha$-carbon C--H bond order parameter.
  }
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
    with the previously published values from DOPS ($T=303$~K, $^2$H NMR, 0.1~M of NaCl) \cite{browning80} and 
    POPC  ($T=300$~K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
    The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
  }
  %\todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
  %We can do this just before submission when citation numbers will not change anymore.}
  %SAMULI: Maybe better to do in revision stage because citations might still change.
\end{figure}

Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
only the absolute values of  $S_{\rm CH}$, we determined the signs of the PS headgroup
 $S_{\rm CH}$ using the S-DROSS experiment \cite{gross97}.
For a given carbon, its S-DROSS dipolar modulation profile in the indirect dimension is a superposition
of sinusoidal functions from the possible orientations of crystallites in the sample (or bilayer patches).
We phase corrected the 2D spectrum in the direct dimension such that positive and negative signs for the $S_{\rm CH}$ 
give rise to profiles that initially increase and decrease, respectively.
In practice, we use the known negative sign of the acyl chain carbons as a reference to perform
the phase correction and interpret the distinct initial slopes of the S-DROSS profiles (Fig. \ref{DPslices}). 
The S-DROSS slice for the $\beta$-carbon clearly shows an initial decrease and therefore its order parameter must be negative.
For the $\alpha$-carbon such analysis is not as trivial due to the two inequivalent order parameters of the two distinct C--H bonds.
However, the beginning of its S-DROSS slice suggests that the larger $S_{\rm CH}$ of the $\alpha$-carbon is positive and the
decrease towards negative values at longer $t_1$ suggests that the smaller  $S_{\rm CH}$ is negative.    
This is confirmed by a SIMPSON simulation
using the  $S_{\rm CH}$ values of $+0.09$ from the dipolar coupling measured here (Fig. \ref{DPslices})
and $-0.02$ from the previous $^2$H\,NMR experiment~\cite{roux91}.
We used the literature value for the smaller  $S_{\rm CH}$ , because the
resolution of our R-PDLF experiment was not sufficient to determine the
magnitude of the small value.
The S-DROSS curve from the SIMPSON simulation with a positive value for the smaller  $S_{\rm CH}$ 
(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
corroborating the interpretation that the smaller  $S_{\rm CH}$  is negative.

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon  $S_{\rm CH}$ is significantly more negative and $\alpha$-carbon
experiences a substantial forking (different  $S_{\rm CH}$ for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but a detailed structrural interpretation is not
available.

We note that the DOPS $^2$H~NMR reference data found in the literature~\cite{browning80,roux90} was obtained by first solvating
the lipids with a buffer solution and then centrifuging the sample to a pellet that was used for the measurements. Such samples have a lower lipid concentration
(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
the gravimetric samples (60~wt~\%) and simulations (approximately 50--60~wt~\%) in this work.
Larger multilamellar repeat distances are expected in the samples with lower lipid
concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
Yet the PS headgroup  $S_{\rm CH}$  measured from gravimetric samples (POPS) in this work
are in good agreement with the results from centrifuged samples~\cite{browning80}. This, together with the rapid decrease of equilibrium repeat distance with addition of monovalent salt~\cite{millman82,rand89}, indicates that the hydration levels of multilamellae are sufficiently similar in the simulations and reference experiments.

>>>>

lipid/counter-ions  & force field & \footnote{Number of lipid molecules}N$_{{\rm l}}$  & \footnote{Number of water molecules}N$_{{\rm w}}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns)  & \footnote{Time used for analyses}t$_{{\rm anal}}$ (ns)  & \footnote{Number of reference for simulation files}files & \tabularnewline

@@

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were calculated by first averaging over time separately
for each lipid in the system, and then calculating the average and
the standard error of the mean over the different lipids. The analysis can be conducted using a
Python program ({\tt calcOrderParameters.py}, available in Ref.~\citenum{MATCHgit}) that uses the
MDAnalysis library \cite{agrawal11,gowers16}.
For united atom models, the positions of hydrogens were generated before the order parameter calculation using the {\tt protonate} tool
of the Gromacs 3 sofware package \cite{gromacsMANUAL}.
The ion number density profiles were calculated using the {\tt gmx density} tool
of the Gromacs sofware package \cite{gromacsMANUAL}.

\subsection{Using the molecular electrometer concept to compare ion binding to negatively charged lipid bilayers 
in simulations and in experiments}

The $S_{\rm CH}$ of the $\alpha$ and $\beta$ carbons in the PC headgroup
decrease proportionally to the amount of positive
charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
and can therefore be used to measure the ion binding affinity.
In addition to ions, the correlation between bound charge and headgroup order parameter change
is empirically observed also for peptides, charged amphiphiles, local anesthetics and charged lipids~\cite{scherer87,beschiaschvili91}.
This concept, known as the molecular electrometer, is especially useful for 
comparison between simulations and experiments, as
the headgroup $S_{\rm CH}$ at varying cation
concentrations can be easily calculated from
simulations~\cite{catte16}. The headgroup $S_{\rm CH}$
of negatively charged PS and PG lipids also exhibit systematic --- although less well understood ---dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, measuring the PC headgroup $S_{\rm CH}$ from 
mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also SI section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing the ion binding to negatively charged membranes.

Calibrating the PC $S_{\rm CH}$ response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer.
This can be done using experimental data from mixtures of PC and
monovalent cationic surfactants (such as POPC and dihexadecyldimethylammonium,
see SI section \ref{electrometerCALIBRATION})~\cite{scherer89,melcr18}.
Additionally, the response of PC headgroup $S_{\rm CH}$ to the negatively
charged PS follows the molecular electrometer in experiments~\cite{scherer87},
which we also quantify here (see SI section \ref{electrometerFORmixtures}).

Studies applying the molecular electrometer have used two different definitions for salt concentration:
The concentrations can be reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids.
In the former case, binding of ions to the lipids leads to a lower bulk concentration than in what was present in the original solvent.
However, the choice of definition has only a marginal effect
to the results in simulations with realistic ion binding affinity
(see SI section \ref{concentrationDEFsection}).
In this work we use the former definition (concentration before solvating the lipids) to be consistent with the reference
experimental data \cite{roux90}.

\section{Results and discussion}

\subsection{Headgroup and glycerol backbone C--H bond order parameters of POPS from $^{13}$C NMR}
The INEPT and 2D R-PDLF experiments from POPS samples gave well resolved spectra for all the
carbons in the headgroup and glycerol backbone regions (Fig. \ref{PShgSIGNSsimpson}).
The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}, whereas
the peaks for $\beta$ and $\alpha$ carbons were assigned according to the
C--H bond order parameters known from previous $^2$H\,NMR experiments~\cite{browning80}.
Slices of the R-PDLF spectra and the resulting  $S_{\rm CH}$ values
are shown in the Supporting Information (Fig. \ref{DPslices}).

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) Refocused-INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental S-DROSS data (points), and SIMPSON simulations with
    the C--H bond order parameter values of $-0.12$ for the $\beta$-carbon and $+0.09$ and $-0.02$
    for the $\alpha$-carbon (blue lines).
    Dashed gray line is the S-DROSS curve from a SIMPSON simulation with a positive value ($+0.02$) 
    for the smaller $\alpha$-carbon C--H bond order parameter.
  }
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS (T = 298 K) measured in this work compared
    with the previously published values from DOPS ($T=303$~K, $^2$H NMR, 0.1~M of NaCl) \cite{browning80} and 
    POPC  ($T=300$~K, $^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work, whereas signs of the PC order parameters are measured previously~\cite{ferreira16}.
    The size of errorbars ($\pm$0.02) shown for $^{13}$C NMR data is justified previously~\cite{botan15,ollila16}. 
  }
  %\todo{Replacing the "literature" with "from Ref X" has been suggested  https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/34.
  %We can do this just before submission when citation numbers will not change anymore.}
  %SAMULI: Maybe better to do in revision stage because citations might still change.
\end{figure}

Since the R-PDLF and previous $^2$H\,NMR experiments \cite{browning80,roux91} give 
only the absolute values of  $S_{\rm CH}$, we determined the signs of the PS headgroup
 $S_{\rm CH}$ using the S-DROSS experiment \cite{gross97}.
For a given carbon, its S-DROSS dipolar modulation profile in the indirect dimension is a superposition
of sinusoidal functions from the possible orientations of crystallites in the sample (or bilayer patches).
We phase corrected the 2D spectrum in the direct dimension such that positive and negative signs for the $S_{\rm CH}$ 
give rise to profiles that initially increase and decrease, respectively.
In practice, we use the known negative sign of the acyl chain carbons as a reference to perform
the phase correction and interpret the distinct initial slopes of the S-DROSS profiles (Fig. \ref{DPslices}). 
The S-DROSS slice for the $\beta$-carbon clearly shows an initial decrease and therefore its order parameter must be negative.
For the $\alpha$-carbon such analysis is not as trivial due to the two inequivalent order parameters of the two distinct C--H bonds.
However, the beginning of its S-DROSS slice suggests that the larger $S_{\rm CH}$ of the $\alpha$-carbon is positive and the
decrease towards negative values at longer $t_1$ suggests that the smaller  $S_{\rm CH}$ is negative.    
This is confirmed by a SIMPSON simulation
using the  $S_{\rm CH}$ values of $+0.09$ from the dipolar coupling measured here (Fig. \ref{DPslices})
and $-0.02$ from the previous $^2$H\,NMR experiment~\cite{roux91}.
We used the literature value for the smaller  $S_{\rm CH}$, because the
resolution of our R-PDLF experiment was not sufficient to determine the
magnitude of the small value.
The S-DROSS curve from a SIMPSON simulation with a positive value for the smaller  $S_{\rm CH}$ 
(dashed grey in Fig. \ref{PShgSIGNSsimpson} (C)) did not agree with the experiment, 
corroborating the interpretation that the smaller  $S_{\rm CH}$ is indeed negative.

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H\,NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon  $S_{\rm CH}$ is significantly more negative, and the $\alpha$-carbon
experiences a substantial forking (different  $S_{\rm CH}$ for the two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but a detailed structrural interpretation is not
available.

We note that the DOPS $^2$H~NMR reference data found in the literature~\cite{browning80,roux90} was obtained by first solvating
the lipids with a buffer solution and then centrifuging the sample to a pellet that was used for the measurements. Such samples have a lower lipid concentration
(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
the gravimetric samples (60~wt~\%) and simulations (approximately 50--60~wt~\%) in this work.
Larger multilamellar repeat distances are expected in the samples with lower lipid
concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
Yet the PS headgroup  $S_{\rm CH}$  measured from gravimetric samples (POPS) in this work
are in good agreement with the results from centrifuged samples~\cite{browning80}. This, together with the rapid decrease of equilibrium repeat distance with addition of monovalent salt~\cite{millman82,rand89}, indicates that the hydration levels of multilamellae are sufficiently similar in the reference experiments and our simulations.
